prolfqua - An R-package for Analysing Proteomics Label Free Quantification Experiments

### 2022-03-28 inst/MyArticle/paper.Rmd
diff

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented (TODO(wew): cite the docu). This makes it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `r BiocStyle::Biocpkg("prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

[TODO SAINTExpress integration, proDA]

@@

Because we used the median as the i-th order statistic $i = \frac{n}{2} + 0.5$. Therefore, $\gamma = i = \frac{n}{2} + 0.5$ and $\delta = n - i + 1 = n - (\frac{n}{2} + 0.5) + 1 = \frac{n}{2} + 0.5 = \gamma$ are used to parameterize the (TODO(wew): what is CDF) CDF of the Beta distribution. We implemented this method in the class `ContrastROPECA` (Figure \@ref(fig:ContrastUML)).

@@

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest AUC scores among all the tested models. By using the beta distribution to model the number of peptides observed, we can further improve the pAUC scores (see Figure TODO(wew)). However, the properties of beta-based probabilities are not well understood; for instance, they have no uniform distribution under the null hypothesis. The FDR estimates obtained when correcting multiple testing with the Benjamini-Hochberg method are strongly biased (see Figure \@ref(fig:FDRfdp) right panel). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. Furthermore, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset (see Figure \@ref(fig:FDRfdp) left panel). To address the issue of missing observation, we could either run `r BiocStyle::Biocpkg("proDA")` or the imputation of group averages method discussed here and afterwards apply `r BiocStyle::Biocpkg("PECA")`.

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model your experiments. Various types of models are available (for an overview see table TODO(wew)), and the contrast specification is consistent and intuitive for all models. The benchmarking results are comparable with existing methods. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for customizing the analysis or adding new features, e.g., generalized linear models (_glm_'s) for missing data modeling or robust linear models (_rlm_'s), quickly.

@@

|       |                                                   |
|:------|:--------------------------------------------------|
|AUC    |Area Under the Curve                               |
|ESI-MS |Electro-Spray-Ionization Mass Spectrometry         |
|FC     |Fold Change                                        |
|$FDP$  |False Discovery Proportion                         |
|$FDR$  |False Discovery Rate                               |
|HPLC   |High-Pressure Liquid Chromatography                |
|LC-MS  |Liquid Chromatography followed by Mass Spectrometry|
|LFQ    |Label-Free Quantification                          |
|LOD    |Limit Of Detection                                 |
|MS     |mass spectrometry                                  |
|OO     |Object-Oriented                                    |
|UML    |Unified Modeling Language                          |

>>>>

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented ([@faraway2016extending], page : ). This makes it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `r BiocStyle::Biocpkg("prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

@@

Because we used the median as the i-th order statistic $i = \frac{n}{2} + 0.5$. Therefore, $\gamma = i = \frac{n}{2} + 0.5$ and $\delta = n - i + 1 = n - (\frac{n}{2} + 0.5) + 1 = \frac{n}{2} + 0.5 = \gamma$ are used to parameterize the cumulative distribution function (CDF) of the Beta distribution. We implemented this method in the class `ContrastROPECA` (Figure \@ref(fig:ContrastUML)).

@@

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest AUC scores among all the tested models. By using the beta distribution to model the number of peptides observed, we can further improve the pAUC scores (see Figure TODO(wew)). However, the properties of beta-based probabilities are not well understood; for instance, they have no uniform distribution under the null hypothesis (TODO remove they). The FDR estimates obtained when correcting multiple testing with the Benjamini-Hochberg method are strongly biased (see Figure \@ref(fig:FDRfdp) right panel). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. Furthermore, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset (see Figure \@ref(fig:FDRfdp) left panel). To address the issue of missing observation, we could either run `r BiocStyle::Biocpkg("proDA")` or the imputation of group averages method discussed here and afterwards apply `r BiocStyle::Biocpkg("PECA")`.

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model your experiments. Various types of models are available (for an overview see table TODO(wew) create table with methods implemented in prolfqua), and the contrast specification is consistent and intuitive for all models. The benchmarking results are comparable with existing methods. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for customizing the analysis or adding new features, e.g., generalized linear models (_glm_'s) for missing data modeling or robust linear models (_rlm_'s), quickly.

@@

|       |                                                   |
|:------|:--------------------------------------------------|
|AUC    |Area Under the Curve                               |
|CDF    |Cumulative Distribution Function                   |
|ESI-MS |Electro-Spray-Ionization Mass Spectrometry         |
|FC     |Fold Change                                        |
|$FDP$  |False Discovery Proportion                         |
|$FDR$  |False Discovery Rate                               |
|HPLC   |High-Pressure Liquid Chromatography                |
|LC-MS  |Liquid Chromatography followed by Mass Spectrometry|
|LFQ    |Label-Free Quantification                          |
|LOD    |Limit Of Detection                                 |
|MS     |mass spectrometry                                  |
|OO     |Object-Oriented                                    |
|UML    |Unified Modeling Language                          |
